---
title: "Peform DB analysis using limma, and compare the results of MACS2 peak calling from different approaches"
author: 
  - name: 'Sepideh Foroutan'
    affiliation: 'Huntington Cancer Immunotherapy Lab'
    url: https://www.monash.edu/discovery-institute/huntington-lab
date: '11-11-2020'
output: html_document
---

We have run MACS2 peak calling on the Cut&Run data, and have analysed it using nextflow, manual shell scripting as well as Galaxy. Here, I perform DB analysis using limma, and additionally, I perform a quick comparison between the results obtained from different approaches. 

```{r}
outPath <- "../output/"
figPath <- "../figure/"
# macs <- read.csv("../macs2_callpeak/ZEB1_macs2_peaks.csv", stringsAsFactors = F)
```

Columns we have in the results are: 

1st: chromosome name
2nd: start position of peak
3rd: end position of peak
4th: name of peak
5th: integer score for display in genome browser (e.g. UCSC)
6th: strand, either "." (=no strand) or "+" or "-"
7th: fold-change
8th: -log10pvalue
9th: -log10qvalue
10th: relative summit position to peak start 
```{r}
library(rtracklayer)
library(magrittr)
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
library(ggupset) 

txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

# peakSet <- import("/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/macs2_callpeak/narrow_peaks/Ikzf3_summits.bed", format = "BED")

file_narrowPeak = "/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/macs2_callpeak/narrow_peaks/Ikzf3_peaks.narrowPeak" 

extraCols_narrowPeak <-
  c(
    signalValue = "numeric",
    pValue = "numeric",
    qValue = "numeric",
    peak = "integer"
  )

gr_narrowPeak <- import(file_narrowPeak, format = "BED",
                        extraCols = extraCols_narrowPeak)

# peakSet <- read.delim("/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/macs2_callpeak/narrow_peaks/Ikzf3_peaks.narrowPeak", header = F)
# # 
# colnames(peakSet) <-
#   c(
#     "chr",
#     "start",
#     "end",
#     "name",
#     "score",
#     "strand",
#     "fold-change",
#     "-log10pvalue",
#     "-log10qvalue",
#     "summit"
#   )
# # 
# # 
# peakSet <- makeGRangesFromDataFrame(
#   df = peakSet,
#   keep.extra.columns = TRUE,
#   ignore.strand = FALSE,
#   seqinfo = NULL,
#   seqnames.field = c(
#     "seqnames",
#     "seqname",
#     "chromosome",
#     "chrom",
#     "chr",
#     "chromosome_name",
#     "seqid"
#   ),
#   start.field = "start",
#   end.field = c("end", "stop"),
#   strand.field = "strand",
#   starts.in.df.are.0based = FALSE
# )

annotatedPeaks <- annotatePeak(
  # peakSet,
  gr_narrowPeak,
  tssRegion = c(-3000, 1000),
  # tssRegion = c(-3000, 3000),
  TxDb = txdb,
  annoDb = "org.Mm.eg.db"
)

pdf(paste0( "../figure/Ikzf3_CnR_peaksNarrow_PieUpset2.pdf"), height = 3.5, width = 6.2)
plotAnnoPie(annotatedPeaks)
upsetplot(annotatedPeaks)
dev.off()

annotated_granges <- as.GRanges(annotatedPeaks)
export(annotated_granges, "../output/annotated_Ikzf3_peak_set_narrow.gtf", format = "GTF")

readr::write_csv(as.data.frame(annotatedPeaks), "../output/annotated_Ikzf3_peak_set_narrow.csv")

```


"verbose =1 : Print immediate results on screen.
genome = Hsapiens : specify the genome.
We also describe two important parameters:
P-Value cutoff: 
- The P-Value cutoff controls the number of binding site in a motif. By default, the P-value cutoff is : 0.0002
- E-Value cutoff: The E-Value cutoff controls the number of motifs to be identified. By default, the E-value cutoff is : 0.0"

```{r}
# order the peaks by qvalue, and take top 250 peaks
peakSet <-  peakSet[order(peakSet$`-log10qvalue`), ]
peakSet2 <-  head(peakSet, n = 500)

gr_narrowPeak2 <- reduce(gr_narrowPeak2)

gr_narrowPeak_resized <- resize(gr_narrowPeak2, width = 50, fix='center')

library(rGADEM)
library(BSgenome.Mmusculus.UCSC.mm10)

mm <-  getSeq(Mmusculus, seqnames(gr_narrowPeak_resized))

gadem <-
  GADEM(
    peakSet,
    verbose = 1,
    genome = Mmusculus,
    seed = 1,
    nmotifs = 10
  )
```


## Count reads to peaks
Get TSS site co-ordinates & count ChIPseq reads to TSS sites:

```{r}
bamPath <- "/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/bam_files/"

# Fetch mouse annotations
# txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
genes <- genes(txdb)
# Get promoter sites and remove extra contigs
tss_sites <- promoters(genes, upstream = 3000, downstream = 1000)
wanted_chr <- seqlevels(tss_sites)[ !grepl(seqlevels(tss_sites), pattern = "_random")]
wanted_chr <- wanted_chr[! grepl(wanted_chr, pattern = "Un_")]
seqlevels(tss_sites, pruning.mode = "coarse") <- wanted_chr
# Write to file
export(tss_sites, "../data/tss_sites.bed", format = "BED")


# Get gene annotations and remove extra contigs
seqlevels(genes, pruning.mode = "coarse") <- wanted_chr
genes <- sort(genes)
# Write to file


export(genes, "../data/mm10_ucsc_genes.gtf", format= "GTF")
export(genes, "../data/mm10_ucsc_genes.bed", format= "BED")

# On the terminal of the RStudio: Format the tss bed files into SAF format
# cut -f4 ./data/tss_sites.bed | paste - ./data/tss_sites.bed | cut -f1,2,3,4,7 > ./data/tss_sites.saf

### RUN the R code instead: Count reads to TSS sites with featureCounts
# featureCounts -T 12 -p -F SAF -a ./data/tss_sites.saf -o ./output/counts_tss.txt /Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/bam_files/AIOLOS_sorted_filtered_mdup.bam

library(Rsubread)

f <- featureCounts(
  files = c(
    paste0(bamPath, "AIOLOS_sorted_filtered_mdup.bam"),
    paste0(bamPath, "NC_sorted_filtered_mdup.bam")
  ),
  annot.ext = "../data/tss_sites.saf",
  isPairedEnd = T,
  nthreads = 12
)

saveRDS(f , paste0(outPath, "Ikzf3_CnR_FeatureCounts_TSS.RDS"))
```

### DB analysis using limma
```{r}
## remove rows that are zero in both:
cc0 <- f$counts
cc <- cc0[rowSums(cc0 > 5) >= 1, ]

# kp <- filterByExpr(cc0, group = c("Ikzf3", "control"))
# cc <- cc0[kp, ]

db <- edgeR::normalizeChIPtoInput(
  input = cc[, "NC_sorted_filtered_mdup.bam"],
  response = cc[, "AIOLOS_sorted_filtered_mdup.bam"],
  plot = T
)

adjp <- p.adjust(p = db$p.value, method = "BH") 
table(db$p.value < 0.05)
table(adjp < 0.05)

adjpData <- data.frame(adjp)
adjpData <- adjpData[order(adjpData$adjp), , drop = F]

adjpData$SYMBOL <-
  mapIds(
    org.Mm.eg.db,
    key = rownames(adjpData),
    keytype = "ENTREZID",
    column = c("SYMBOL")
  )

adjpData$ENTREZID <- rownames(adjpData)

write.table(adjpData, paste0(outPath, "Ikzf3_CnR_FeatureCount_edgeR_DB_stats_FiltCount5.txt"),
            row.names = F, sep = "\t")

# topDB <- head(adjpData, 1400)
topDB <- adjpData[adjpData$adjp < 0.05, ]

length(intersect(topDB$SYMBOL, sep$SYMBOL))    ## 461
length(intersect(topDB$SYMBOL, csaw3$symbol))  ## 240

```


## Compare with Nextflow
```{r}

file_narrowPeak_nf = "/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/nexflow/ikfz3_results/bwa/mergedLibrary/macs/narrowPeak/Ikfz3_IP_R1_peaks.narrowPeak" 
extraCols_narrowPeak_nf <- c(signalValue = "numeric", pValue = "numeric",
                          qValue = "numeric", peak = "integer")

gr_narrowPeak_nf <- import(file_narrowPeak_nf, format = "BED",
                        extraCols = extraCols_narrowPeak_nf)

seqlevels(gr_narrowPeak_nf) <- paste0("chr", seqlevels(gr_narrowPeak_nf))
seqlevels(gr_narrowPeak_nf)[seqlevels(gr_narrowPeak_nf) == "chrMT" ] <- "chrM"

annotatedPeaks_nf <- annotatePeak(
  gr_narrowPeak_nf,
  tssRegion = c(-3000, 1000),
  TxDb = txdb,
  annoDb = "org.Mm.eg.db"
)


pdf(
  paste0("../figure/Ikzf3_CnR_peaksNarrow_PieUpset_nf.pdf"),
  height = 3.5,
  width = 6.2
)
plotAnnoPie(annotatedPeaks_nf)
upsetplot(annotatedPeaks_nf)
dev.off()

annotated_granges_nf <- as.GRanges(annotatedPeaks_nf)
# annotated_granges <- as.GRanges(annotatedPeaks)
## 54 pairs
readr::write_csv(as.data.frame(annotatedPeaks_nf), "../output/annotated_Ikzf3_peak_set_narrow_nf.csv")


```

## Compare with Galaxy

```{r}
# peakSetSeb <- import("/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/bam_galaxy_Seb/Galaxy38-[MACS2_callpeak_AIOLOS].bed")
# s <- read.delim("/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/bam_galaxy_Seb/Galaxy38-[MACS2_callpeak_AIOLOS].bed", header = F)

# peakSetSeb <- import("/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/bam_galaxy_Seb/Galaxy_Ikzf3.bed", format = "BED")

file_narrowPeakSeb = "/Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/bam_galaxy_Seb/Galaxy38-[MACS2_callpeak_AIOLOS].bed"

extraCols_narrowPeakSeb <- c(signalValue = "numeric", pValue = "numeric",
                          qValue = "numeric", peak = "integer")

gr_narrowPeakSeb <- import(file_narrowPeakSeb, format = "BED",
                        extraCols = extraCols_narrowPeakSeb)

# On the terminal of the RStudio: Format the tss bed files into SAF format
# cut -f1,2,3,4,8 /Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/bam_galaxy_Seb/Galaxy38-[MACS2_callpeak_AIOLOS].bed > /Users/mfor0011/Documents/data/Ikzf/Ikfz3_CnR/bam_galaxy_Seb/Galaxy_Ikzf3.bed

annotatedPeaksSeb <- annotatePeak(
  gr_narrowPeakSeb,
  tssRegion = c(-3000, 1000),
  TxDb = txdb,
  annoDb = "org.Mm.eg.db"
)

pdf(paste0( "../figure/Ikzf3_CnR_peaksGalaxy_PieUpset.pdf"), height = 3.5, width = 6.2)
plotAnnoPie(annotatedPeaksSeb)
upsetplot(annotatedPeaksSeb)
dev.off()

annotated_grangesSeb <- as.GRanges(annotatedPeaksSeb)
export(annotated_grangesSeb, "../output/annotated_Ikzf3_peak_set_Galaxy.gtf", format = "GTF")

readr::write_csv(as.data.frame(annotatedPeaksSeb), "../output/annotated_Ikzf3_peak_set_Galaxy.csv")
```

1202 common genes
```{r}
## 1404
sep <- as.data.frame(annotatedPeaks)
## 1398
gal <- as.data.frame(annotatedPeaksSeb)

nf <- as.data.frame(annotatedPeaks_nf)

length(intersect(sep$SYMBOL, gal$SYMBOL)) ## 1202
length(intersect(sep$SYMBOL, nf$SYMBOL))  ## 1138
length(intersect(gal$SYMBOL, nf$SYMBOL))  ## 1115
```


## Coverage of bams
```{r}
#load library
library(Rsamtools)

#read in entire BAM file
bam <- scanBam(paste0(bamPath, "AIOLOS_sorted_filtered_mdup.bam"))

#names of the BAM fields
names(bam[[1]])
# [1] "qname"  "flag"   "rname"  "strand" "pos"    "qwidth" "mapq"   "cigar"
# [9] "mrnm"   "mpos"   "isize"  "seq"    "qual"

#distribution of BAM flags
table(bam[[1]]$flag)
#     83     99    147    163 
# 841778 834539 834525 841801 

#      0       4      16 
#1472261  775200 1652949

#function for collapsing the list of lists into a single list
#as per the Rsamtools vignette
.unlist <- function (x){
   ## do.call(c, …) coerces factor to integer, which is undesired
   x1 <- x[[1L]]
   if (is.factor(x1)){
      structure(unlist(x), class = "factor", levels = levels(x1))
   } else {
      do.call(c, x)
   }
}

#store names of BAM fields
bam_field <- names(bam[[1]])

#go through each BAM field and unlist
list <- lapply(bam_field, function(y) .unlist(lapply(bam, "[[", y)))

#store as data frame
bam_df <- do.call("DataFrame", list)
names(bam_df) <- bam_field

dim(bam_df)
# 3352643      13

# saveRDS(bam_df, paste0(outPath, "Ikzf3_CnR_bam_DataFrame.RDS"))
```

```{r}
#use chr11 as an example
#how many entries on the negative strand of chr11?
table(bam_df$rname == 'chr11')
# FALSE    TRUE 
#3875997   24413

#function for checking negative strand
check_neg <- function(x){
  if (intToBits(x)[5] == 1){
    return(T)
  } else {
    return(F)
  }
}

#test neg function with subset of chr11
test <- subset(bam_df, rname == 'chr11')
dim(test)
#[1]178108     13
table(apply(as.data.frame(test$flag), 1, check_neg))
#number same as above
#FALSE  TRUE 
# 89053 89055 

#function for checking positive strand
check_pos <- function(x){
  if (intToBits(x)[3] == 1){
    return(F)
  } else if (intToBits(x)[5] != 1){
    return(T)
  } else {
    return(F)
  }
}

#check pos function
table(apply(as.data.frame(test$flag), 1, check_pos))
#looks OK
#FALSE  TRUE 
# 89055 89053 

# store the mapped positions on the plus and minus strands
chr11_neg <- bam_df[bam_df$rname == 'chr11' &
                    apply(as.data.frame(bam_df$flag), 1, check_neg),
                    'pos'
                   ]
length(chr11_neg)
#[1] 89055
chr11_pos <- bam_df[bam_df$rname == 'chr11' &
                    apply(as.data.frame(bam_df$flag), 1, check_pos),
                    'pos'
                   ]
length(chr11_pos)
#[1] 32013
 
#calculate the densities
chr11_neg_density <- density(chr11_neg)
chr11_pos_density <- density(chr11_pos)
 
#display the negative strand with negative values
chr11_neg_density$y <- chr11_neg_density$y * -1
 
plot(chr11_pos_density,
     ylim = range(c(chr11_neg_density$y, chr11_pos_density$y)),
     main = "Coverage plot of Cut & Run reads for Chr 11",
     xlab = "Chromosome 11",
     col = 'blue',
     lwd=2.5)
lines(chr11_neg_density, lwd=2.5, col = 'red')

#say I am interested in region 31480000 to 39080000

lower <- 98464902
upper <- 98546031
 
chr11_pos_interest <- chr11_pos[chr11_pos > lower & chr11_pos < upper]
#check how many entries we have
length(chr11_pos_interest)
#[1] 16220
 
#do the same for the negative strand
chr11_neg_interest <- chr11_neg[chr11_neg > lower & chr11_neg < upper]
length(chr11_neg_interest)
#[1] 7744
 
#now continue with the code above
#but with our two new vectors of interest
chr11_neg_density <- density(chr11_neg_interest)
chr11_pos_density <- density(chr11_pos_interest)
  
#display the negative strand with negative values
chr11_neg_density$y <- chr11_neg_density$y * -1
  
plot(chr11_pos_density,
     ylim = range(c(chr11_neg_density$y, chr11_pos_density$y)),
     main = "Coverage plot",
     xlab = "Chromosome 11, Ikzf3",
     col = 'blue',
     lwd=2.5,
     type='h'
)
lines(chr11_neg_density, lwd=2.5, col = 'red', type='h')
```

